\thetitle

### 2021-05-19 writing/discussion.tex
local edits here and there

At first glance, these measurements seem to contradict the results
of~\citet{vss-popl-2017}. They report an average slowdown of 6.2x and a
worst-case of 17.2x on the fully-typed Python benchmarks in Reticulated
Python with the blame map enabled.  Unfortunately, the orders-of-magnitude
higher cost of Shallow Racket seems closer to the truth.  There are at
least four broad issues that skew the earlier results:
\begin{enumerate}

@@

\item While Reticulated attempts to infer types for local variables,
  often local variables end up with type
{\tt Dynamic} due the impoverished nature of the type system,  which does
    not allow the ascription of precise
    types~\citet{g-dissertation-2020}. Code with the type {\tt
Dynamic} has fewer constraints to check at run-time---and much less information
to track in the blame map.

\item \citet{vss-popl-2017} use small benchmarks.  Four have since been retired
from the official Python benchmark suite because they are too small,
unrealistic, and unstable.\footnote{Release notes:
\url{https://pyperformance.readthedocs.io/changelog.html}}

\item Reticulated Python does not seem to employ any crucial
  implementation techniques that can mitigate the significant slowdown due to
    the blame map observed in Shallow Racket. For example Reticulated Python  runs the simplest
    benchmark from the GTP suite (translated into Python) in approximately
    40 seconds without blame but times out after 10 minutes with blame.

>>>>

At first glance, these measurements seem to contradict the results
of~\citet{vss-popl-2017}. They report an average slowdown of 6.2x and a
worst-case slowdown of 17.2x on the fully-typed Python benchmarks in Reticulated
Python when the blame map gets enabled.  Unfortunately, the average slowdown of 
133x and the
worst-case slowdown of 560x  due to to blame  in Shallow Racket seems closer to the truth.  There are at
least three broad issues that skew the earlier results:
\begin{enumerate}

@@

\item While Reticulated attempts to infer types for local variables, the impoverished nature of its type
    system does not allow the ascription of precise types and often resorts
    to type {\tt Dynamic}~\citep[section~5.4.4]{g-dissertation-2020}. 
    Code with the type {\tt Dynamic}
    has fewer constraints to check at run-time---and much less information
    to track in the blame map.

\item \citet{vss-popl-2017} use small benchmarks.  Four have since been retired
from the official Python benchmark suite because they are too small,
unrealistic, and unstable.\footnote{Release notes:
\url{https://pyperformance.readthedocs.io/changelog.html}}  On the flip
    side, all the benchmarks in the GTP suite are larger than the official Python
    benchmarks. Reticulated Python  runs the
    translation of the smallest GTP benchmark in
    approximately 40 seconds without blame but times out after 10 minutes
    with blame.

### 2021-05-19 writing/introduction.tex
local edits here and there

The paper's {\it second contribution\/} is a set of {\em results from
applying the evaluation method to three distinct blame assignment
strategies in approximately 72,000 different scenarios\/} (see
sec.~\ref{sec:results}): (1)~{\it Transient\/}, i.e. Reticulated's
tracking of typed/untyped boundaries~\citep{vss-popl-2017}; (2)~{\it
Natural\/}, i.e. Typed Racket's use of a higher-order contract system and
its blame assignment~\citep{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e.
the approach of industrial systems, which forgo blame in favor of error
messages from the safety checks of the underlying language. The results
(see sec.~\ref{sec:discussion}) are at least somewhat surprising.  In
principle they validate the conjectures behind the work of theoreticians.
\mf{don't we also want to say the run-time checks alone help?}
A blame assignment strategy helps with the search for
impedance mismatches between the specified types and the behavior of
untyped components.  Natural's wrapper-based blame tracking is
more useful than Transient's ``collective blame'' tracking algorithm,
which in turn is superior to Erasure. {\em But\/}, the application of the
method also indicates problems with the expectations of theoreticians. 
In contrast with the theoretical differences between the methods,
{\it Natural\/} is only marginally more useful than {\it Transient\/}, and neither of the two
blame assignment methods are highly superior to {\it Erasure\/}. Additionally, 
the cost of Transient's blame can be huge.  In turn, these
problems suggest that, on one hand, the existing theory does not properly
predict the behavior of blame in real systems with real programs,
and on the other hand, the existing practice lacks data and alternative
experiments to assess the entire landscape of blame assignment.

>>>>

The paper's {\it second contribution\/} is a set of {\em results from
applying the evaluation method to three distinct checking regimes for
gradual types and their blame assignment
strategies in approximately 72,000 different scenarios\/} (see
sec.~\ref{sec:results}): (1)~{\it Transient\/}, i.e. Reticulated's
inlined type assertions and collaborative tracking of typed/untyped boundaries~\citep{vss-popl-2017}; (2)~{\it
Natural\/}, i.e. Typed Racket's use of a higher-order contract system and
its blame assignment~\citep{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e.
the approach of industrial systems, which forgo type checks and blame in favor of error
messages from the safety checks of the underlying language. The results
(see sec.~\ref{sec:discussion}) are at least somewhat surprising.  In
principle they validate the conjectures behind the work of theoreticians.
Type checks and blame work together to help with the search for
impedance mismatches between the specified types and the behavior of
untyped components.  Natural's wrapper-based type checks and blame tracking are
more useful than Transient's type in-lined assertions and ``collective blame'' tracking algorithm,
which in turn are superior to Erasure. {\em But\/}, the application of the
method also indicates problems with the expectations of theoreticians. 
In contrast with the theoretical differences between the methods,
{\it Natural\/} is only marginally more useful than {\it Transient\/}, and neither of the two
checking and blame assignment methods are highly superior to {\it Erasure\/}. Additionally, 
the cost of Transient's blame can be huge.  In turn, these
problems suggest that, on one hand, the existing theory does not properly
predict the behavior of blame in real systems with real programs,
and on the other hand, the existing practice lacks data and alternative
experiments to assess the entire landscape of the pragmatics of blame.

### 2021-05-19 writing/paper.tex
Fix citation style
Add blame trail example to discussion, in new subsection
authors
turn on ACM thingies
improve prose, mark up 2 questionable phrases
local edits here and there

\documentclass[acmsmall,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}

@@

\author{Nobody In Particular}
% \affiliation{\institution{PLT} \city{Evanston} \country{Illinois} }
\affiliation{\relax} % \institution{PLT} \city{Evanston} \country{Illinois} }
% \email{chrdim@northwestern.edu}          %% \email is recommended

@@

\usepackage{upgreek}

@@

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Programming language theoreticians develop blame assignment systems
and prove blame theorems for gradually typed programming languages.
Practical implementations of gradual typing almost completely ignore
the idea of blame assignment.  This contrast raises the question
whether blame assignment provides any value to the working programmer
and poses the challenge of how to evaluate the effectiveness of blame
assignment strategies. This paper contributes (1) the first evaluation
method for blame assignment strategies and (2) the results from applying
it to three different gradually typed languages. These results confirm that blame assignment
is helpful to a rationally acting programmer.
\end{abstract}

>>>>

\documentclass[acmsmall]{acmart}
% \settopmatter{printfolios=true,printccs=false,printacmref=false}

\citestyle{acmauthoryear}

@@

\author{Lukas Lazarek}
\affiliation{%
  \institution{PLT @ Northwestern University}
  \city{Evanston}
  \state{Illinois}
  \country{USA}
}
\email{lukas.lazarek@eecs.northwestern.edu}

\author{Ben Greenman}
\orcid{0000-0001-7078-9287}
\affiliation{%
  \institution{PLT @ Northeastern University}
  \city{Boston}
  \state{Massachusetts}
  \country{USA}
}
\email{benjaminlgreenman@gmail.com}

\author{Matthias Felleisen}
\orcid{0000-0001-6678-1004}
\affiliation{%
  \institution{PLT @ Northeastern University}
  \city{Boston}
  \state{Massachusetts}
  \country{USA}
}
\email{matthias@ccs.neu.edu}

\author{Christos Dimoulas}
\orcid{0000-0002-9338-7034}
\affiliation{%
  \institution{PLT @ Northwestern University}
  \city{Evanston}
  \state{Illinois}
  \country{USA}
}
\email{chrdimo@northwestern.edu}
\renewcommand{\shortauthors}{Lazarek, Greenmanm Felleisen, Dimoulas}

@@

\usepackage{upgreek}
\usepackage{multirow}

@@

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
%% \setcopyright{none}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Programming language theoreticians develop blame assignment systems and
  prove blame theorems for gradually typed programming languages.
  Practical implementations of gradual typing almost completely ignore the
  idea of blame assignment.  This contrast raises the question whether
  blame provides any value to the working programmer and poses
  the challenge of how to evaluate the effectiveness of blame assignment
  strategies. This paper contributes (1) the first evaluation method for
  blame assignment strategies and (2) the results from applying it to
  three different semantics for gradual typing. These results cast doubt to
  the theoretical effectiveness of blame. In most scenarios,  strategies
  with imprecise blame assignment are as helpful to a rationally acting
  programmer as strategies with provably correct blame.
\end{abstract}

### 2021-05-19 writing/related.tex
editng discussion about reticulated
local edits here and there

The literature on gradual typing presents many semantics beyond the three used
here and three additional blame strategies.  Pyret (pyret.org) assigns fixed-size
data types the Natural semantics and functions a Transient semantics. The
Forgetful~\citep{cl-icfp-2017} and the Amnesic~\citep{gfd-oopsla-2019} semantics
are similar to Transient but use wrappers instead of in-lined checks.
Nom~\citep{mt-oopsla-2017} and other \emph{concrete\/}
semantics~\citep{wnlov-popl-2010, rsfbv-popl-2015, rzv-ecoop-2015,
rat-oopsla-2017} assume that every value comes with a type tag and use tag
checks to supervise the interactions between typed and untyped code.  The semantics~\citep{tlt-popl-2019,
etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the Abstracting
Gradual Typing technique~\citep{gct-popl-2016} are variants of Natural.
 Monotonic References~\citep{svctg-esop-2015, rsfbv-popl-2015, sfrbcsb-popl-2014, kas-pldi-2019} 
differ from Natural in the treatment of mutable data by
associating heap locations with a type and permitting only updates that 
make that type more precise.  
Among these representative semantics, only Amnesic, Nom and Monotonic References present innovative blame
strategies.  The experiment excludes the first because it is
merely a theoretical construction, the second because it imposes severe restrictions on
programmers and the third because it would require a complete
re-engineering of the Racket runtime. That said, the paper offers a
detailed chart that the creators of these blame strategies can transplant to their
linguistic setting and determine the value of their blame strategies.

>>>>

The literature on gradual typing presents many semantics beyond the three used
here and three additional blame strategies.  Pyret (pyret.org) assigns fixed-size
data types the Natural semantics and functions a Transient semantics. The
Forgetful~\citep{cl-icfp-2017} and the Amnesic~\citep{gfd-oopsla-2019} semantics
are similar to Transient but use wrappers instead of in-lined checks.
Nom~\citep{mt-oopsla-2017} and other \emph{concrete\/}
semantics~\citep{wnlov-popl-2010, rsfbv-popl-2015, rzv-ecoop-2015,
rat-oopsla-2017} assume that every value comes with a type tag and use tag
checks to supervise the interactions between typed and untyped code.  The semantics~\citep{tlt-popl-2019,
etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the Abstracting
Gradual Typing technique~\citep{gct-popl-2016} are variants of Natural.
The Monotonic semantics~\citep{svctg-esop-2015, rsfbv-popl-2015, sfrbcsb-popl-2014, kas-pldi-2019} 
differs from Natural in the treatment of mutable data by
associating heap locations with a type and permitting only updates that 
do not make the type less precise.  
Among these representative semantics, only Amnesic, Nom, and Monotonic present innovative blame
strategies.  The experiment excludes the first because it is
merely a theoretical construction, the second because it imposes severe restrictions on
programmers, and the third because it would require a complete
re-engineering of the Racket runtime. 
%That said, the paper offers a
%detailed chart that the creators of these blame strategies can transplant to their
%linguistic setting and determine the value of their blame strategies.
